---
title: "milestone2"
author: "Nilson Palma, Michelle Zhen Huang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Milestone 2

This is a team assignment; each team should complete and turn in an html document created from an Rmd or Qmd via Github. Please include code and output for the following components:

##Description of dataset:

What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
How does the dataset relate to the group problem statement and question?
##

Import statements for all datasets:
##

NOTE: Please use datasets available in the PHW251 Project Data github repoLinks to an external site. (this is important to make sure everyone is using the same datasets)
Use appropriate import function and package based on the type of file
Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)

##Document the import process
```{r document}


```

##Clean up column names (recommend using snake case, which is lower case text with underscores between words)

```{r cleanup}
plot(pressure)
```

##Identify data types for 5+ data elements/columns/variables:
##

##Identify 5+ data elements required for your specified scenario. If <5 elements are required to complete the analysis, please choose additional variables of interest in the data set to explore in this milestone.
##

##Utilize functions or resources in RStudio to determine the types of each data element (i.e. character, numeric, factor)
##Identify the desired type/format for each variable—will you need to convert any columns to numeric or another type? 
##Provide a basic description of the 5+ data elements:

Numeric: mean, median, range
Character: unique values/categories
Or any other descriptives that will be useful to the analysis
##Describe cleaning that each data source may need:

NOTE: There is no requirement for any data cleaning in this milestone, but there may be cleaning you can do when you import your data! Please list out the anticipated data cleaning needed. 
Examples:
Data elements that need to be converted to a different type
Data elements that need cleaning or re-categorization
Data elements that may be used for future joins (i.e. state, county)

