title: "Milestone 2"
author: "Nilson Palma, Michelle Zhen Huang"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

```

# Scenario 1: Infectious disease outbreak (simulated) in California
The sim_novelid_CA.csv file and sim_novelid_LACounty.csv file together represent simulated morbidity for the entire state of California. While it’s certainly possible for data from different sources to adhere to a format standard, one of the challenges of working with secondary data is that it often does not work out that way. Therefore to prepare each dataset (milestone #3) so that they can be combined into one, whole state, dataset you will need to do the following:

Morbidity datasets (from CA and LA County)
Recode column names, values, or formats that are in discordant (dates, etc)
Combine morbidity datasets into a single dataset
Select demographic and geographic strata (s) of interest
Aggregate the data into a new dataframe to  include only one row per strata of interest
Population dataset
Recode values to be consistent with morbidity datasets so they can be joined
Select demographic or geogrphic strata(s) of interest
Create a rate metric and summarize to include only one row per strata of interest


###For all scenarios, please turn in an html document created from an Rmd or Qmd with the following components:
Subset rows or columns, as needed
Create new variables needed for analysis (minimum 2)
New variables should be created based on existing columns; for example
Calculating a rate
Combining character strings
Aggregation
Clean variables needed for analysis (minimum 2)
#Examples:
Recode invalid values
Handle missing data
Recode categories
Data re-structured as needed (aggregated/summarized and/or pivoted)
Data dictionary based on clean dataset (minimum 4 data elements), including:
Variable name
Data type
Description
One or more tables with descriptive statistics for 4 data elements
Html output that is professionally prepared for presentation
Only the necessary information is outputted (you should suppress, for example, entire data frame outputs)
Use of headers and sub headers to create an organized document


```{r setup_lib}

library(janitor)
library(tidyverse)
library(dplyr)
library(readr)
install.packages("aweek")
library(aweek)
library(kableExtra)

```



```{r cleandata}


getwd()
setwd("../outbreak_data")

ca_pop <- read.csv("ca_pop_2023.csv")
sim_novelid_ca <- read.csv("sim_novelid_CA.csv")
sim_novelid_la <- read.csv("sim_novelid_LACounty.csv")

sim_novelid_ca <- sim_novelid_ca %>% clean_names()
sim_novelid_la <- sim_novelid_la %>% clean_names()

# Standardize naming and select for columns
simca1 <- sim_novelid_ca %>% 
  select(
    county,
    age_cat,
    sex,
    race_eth = race_ethnicity,
    time_int,
    new_infections,
    cumulative_infected,
    new_severe,
    cumulative_severe
  )
simla1 <- sim_novelid_la %>% 
  mutate(county = "Los Angeles") %>% 
  select(
    county,
    age_cat = age_category,
    sex,
    race_eth,
    dt_diagnosis = dt_dx,
    new_infections = dx_new,
    cumulative_infected = infected_cumulative,
    new_severe = severe_new,
    cumulative_severe = severe_cumulative
  )

simca1 <- simca1 %>% drop_na()
simla1 <- simla1 %>% drop_na()


str(simca1)
str(simla1)

simla1 <- simla1 %>%
  mutate(dt_diagnosis = dmy(dt_diagnosis))

?week


set_week_start(7) 
# CDC Has a Sunday start date
## https://www.epirhandbook.com/en/new_pages/dates.html#dates_epi_wks

simca2 <- simca1 %>%
  mutate(
    year = substr(time_int, 1, 4),   # Extract 2023
    week = substr(time_int, 5, 6),    # Extract 22–52
    cdc_week = aweek::get_date(week = week, year = year)
  ) %>%
  select(-year, -week)

simla2 <- simla1 %>% 
  mutate(
    cdc_week = get_date(
      week = isoweek(dt_diagnosis),
      year = year(dt_diagnosis)
    )
  )

# Convert race numeric to named factor
simca3 <- simca2 %>%
  mutate(
    race_eth = recode_factor(race_eth,
                             `1` = "White, Non-Hispanic",
                             `2` = "Black, Non-Hispanic",
                             `3` = "American Indian or Alaska Native, Non-Hispanic",
                             `4` = "Asian, Non-Hispanic",
                             `5` = "Native Hawaiian or Pacific Islander, Non-Hispanic",
                             `6` = "Multiracial (two or more of above races), Non-Hispanic",
                             `7` = "Hispanic (any race)",
                             `9` = "Unknown"
    )
  ) %>% 
  select(-time_int)

# Convert race (already character) to factor
simla3 <- simla2 %>%
  mutate(race_eth = factor(race_eth))

#Convert dt_diagnosis which is chr to date to merge (NEW)
str(simca3$dt_diagnosis)
simca3$dt_diagnosis <- as.Date(simca3$dt_diagnosis)


sim_all <- bind_rows(simca3, simla3)

```


```{r part2}

str(ca_pop1)

# Standardize naming and select for columns
ca_pop1 <- ca_pop %>% 
  select(
    county,
    age_cat,
    sex,
    race_eth = race7,
    pop
  )

unique(ca_pop1$race_eth)

#Standardize race_eth
ca_pop2<- ca_pop1 %>%
  mutate(
    race_eth = recode_factor(race_eth,
                             "WhiteTE NH" = "White, Non-Hispanic",
                             "Black NH" = "Black, Non-Hispanic",
                             "AIAN NH" = "American Indian or Alaska Native, Non-Hispanic",
                             "Asian NH" = "Asian, Non-Hispanic",
                             "NHPI NH" = "Native Hawaiian or Pacific Islander, Non-Hispanic",
                             "MR NH" = "Multiracial (two or more of above races), Non-Hispanic",
                             "Hispanic" = "Hispanic (any race)"
    )
  ) 


#reclassify age_cat to 0-17 to match sim_novelid_ca dataset, and aggregate population data

ca_pop3 <- ca_pop2 %>%
  mutate(age_cat = case_when(
    age_cat %in% c("0-4", "5-11", "12-17") ~ "0-17",
    TRUE ~ age_cat
  )) %>%
  group_by(county, age_cat, sex, race_eth) %>%
  summarise(pop = sum(pop, na.rm = TRUE)) %>% 
  ungroup()

#Remove "county" in county column to have data sets match for County column
sim_all1 <- sim_all %>%
  mutate(
    county = str_to_title(str_remove(county, " County$")),  # remove trailing " County"
    county = str_trim(county)
  )

ca_pop4 <- ca_pop3 %>%
  mutate(
    county = str_to_title(str_trim(county))
  )

#combine sim_all with ca_pop data set
sim_pop <- sim_all1 %>%
  left_join(ca_pop4, 
            by = c("county", "age_cat", "sex", "race_eth"))

sim_pop_clean <- sim_pop %>% 
  select(
    county,
    age_cat,
    sex,
    race_eth,
    new_infections,
    cumulative_infected,
    new_severe,
    cumulative_severe,
    pop
  )


#combining rows (with dt_diagnosis and cdc_week not included):
sim_combined <- sim_pop_clean %>%
  group_by(county, age_cat, sex, race_eth) %>%
  summarise(
    new_infections = sum(new_infections, na.rm = TRUE),
    cumulative_infected = sum(cumulative_infected, na.rm = TRUE),
    new_severe = sum(new_severe, na.rm=TRUE),
    cumulative_severe = sum(cumulative_severe, na.rm =TRUE),
    pop = sum(pop, na.rm = TRUE)
  ) %>% 
  ungroup()

```
```{r part 3}

#create new variables for proportions and prevalence of both infected and severe
sim_combined_var <- sim_combined %>%
  mutate(
    incidence_prop = new_infections / pop,
    prevalence_infected = cumulative_infected / pop,
    inc_severe_prop = new_severe / pop,
    prev_severe = cumulative_severe /pop
  )


#summary table
summary_table <- sim_combined_var %>%
  summarise(
    `Total Infections` = sum(new_infections, na.rm = TRUE),
    `Mean New Infections` = mean(new_infections, na.rm = TRUE),
    `Total Severe Cases` = sum(new_severe, na.rm = TRUE),
    `Mean Cumulative Infected` = mean(cumulative_infected, na.rm = TRUE),
  )

kable(summary_table, 
      caption = "Summary of Infection Metrics",
      digits = 2) 

```

